State of the Practice for Medical Imaging Software

### 2022-04-04 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Complete draft of edits to Section 6 (Comparison of Artifacts)

As part of filling in the measurement template the MI packages were examined for
the presence of artifacts, which were then categorized by frequency.
Table~\ref{artifactspresent} groups them into categories of common (20 to 29
($>$67\%) packages), uncommon (10 to 19 (33-67\%) packages), and rare (1 to 9
($<$33\%) packages). The full measurements are summarized in
\citet{Dong2021-Data}.  The details on which projects use which types of
artifacts are summarized in Tables~\ref{tab_maintainability_docs}
and~\ref{tab_Visibility/Transparency_docs} for documents related to
maintainability and visibility, respectively.

\begin{table}[ht!]
    \begin{center}
    \begin{tabular}{ p{4.6 cm} p{5.6 cm} p{5 cm}}
    \toprule
    \textbf{Common} & \textbf{Uncommon} & \textbf{Rare} \\
    \midrule
    README (29) & Build scripts (18) & Getting Started (9)\\
    Version control (29) & Tutorials (18) & Developer's manual (8)\\
    License (28) & Installation guide (16) & Contributing (8)\\
    Issue tracker (28) & Test cases (15) & API documentation (7)\\
    Change request (28) & Authors (14) & Dependency List (7)\\
    User manual (22) & Frequently Asked Questions (FAQ) (14) & Troubleshooting guide (6)\\
    Release info. (22) & Acknowledgements (12) & Product roadmap (5)\\
     & Changelog (12) & Design Documentation (5)\\
     & Citation (11) & Code Style Guide (3)\\
     & & Code of Conduct (1)\\
    \bottomrule
    \end{tabular}
    \caption{Artifacts Present in MI Packages, Classified by Frequency (The number in brackets is the number of occurrences)}
    \label{artifactspresent}
    \end{center}
\end{table}

We answer \rqref{RQ_CompareArtifacts} by comparing the artifacts that we
observed in MI repositories to those observed and recommended for research
software in general. Our comparison may point out areas where some MI software
packages fall short of current best practices. This is not intended to be a
criticism of any existing packages, especially since in practice not every
project needs to achieve the highest possible quality. However, rather than
delve into the nuances of which software can justify compromising which
practices we will write our comparison under the ideal assumption that every
project has sufficient resources to match best practices.

Table~\ref{Tbl_Guidelines} shows that the MI artifacts generally match the recommendations found in nine current research software development guidelines:
\begin{inparaenum}[i)]
\item United States Geological Survey Software Planning Checklist
\citep{USGS2019},
\item DLR (German Aerospace Centre) Software Engineering Guidelines
\citep{TobiasEtAl2018}, 
\item Scottish Covid-19 Response Consortium Software Checklist
\citep{BrettEtAl2021},
\item Good Enough Practices in Scientific Computing \citep{WilsonEtAl2016},
\item xSDK (Extreme-scale Scientific Software Development Kit) Community Package
Policies \citep{SmithAndRoscoe2018},
\item Trilinos Developers Guide \citep{HerouxEtAl2008},
\item EURISE (European Research Infrastructure Software Engineers') Network
Technical Reference \citep{ThielEtAl2020},
\item CLARIAH (Common Lab Research Infrastructure for the Arts and Humanities)
Guidelines for Software Quality \citep{vanGompelEtAl2016}, and
\item A Set of Common Software Quality Assurance Baseline Criteria for Research
Projects \citep{OrvizEtAl2017}.
\end{inparaenum}
As examples of compliance with most guidelines, MI developers commonly include
the almost ubiquitous advice to have LICENSE files and README files. , although,
as for other software, the content of README files is not standardized
\citep{PranaEtAl2018}. 15 of 26 artifacts (58\%).  Although in the rare column, some MI developers include
an FAQ (Frequently Asked Questions) document as recommended by
\citep{OrvizEtAl2017}, \citep{ThielEtAl2020} and \citep{vanGompelEtAl2016}. MI
developers also follow the ubiquitous recommendation to have a README file
Compared to ocean modelling software \citep{JungEtAl2022}, MI software pays more
attention to artifacts related to testing.

@@

A gap exists in scientific
computing development practices and software engineering recommendations
\citep{OwojaiyeEtAl2021_CSE}.

Three of the items that appear in Table~\ref{artifactspresent} do not explicitly
appear in the software development guidelines: i) list of related publications,
ii) performance information and iii) video guides.  The occurrence of the first
two items in this list are likely more a consequence of how we collected our
list of MI artifacts, rather than being a unique practice of MI developers. We
highlighted publications and performance information because we were explicitly
looking for both as part of filling in our measurement template. Related
publications and performance information certainly appear in other research
software projects; the practices just aren't highlighted in guidelines because
they are implicit in other documentation recommendations, such as theory
documentation and user guides. The observation of a video guide artifact is more
novel.  \citet{Fogel2005} recommends videos for open source projects, and other
projects likely use them, but the practice has apparently not yet made it into
research software community's guidelines.  Putting aside the debate of the
effectiveness of learning from print versus video, video documentation is
growing in popularity, with a majority of Generation Z learners preferring video
to print \citep{Genota2018}.

Although the MI community participates in most of the practices we found listed
in the general research software guidelines, some recommended practices were not
observed or rarely observed. For instance API documentation was rarely observed
for MI software, but it is a frequently recommended artifact
\citep{SmithAndRoscoe2018, ThielEtAl2020, vanGompelEtAl2016, OrvizEtAl2017,
SSI2022, Zadka2018}.  In addition to the items in the last column of
Table~\ref{artifactspresent}, we can add the following community recommended
items that we rarely, if ever, observed:

\begin{itemize}
\item A roadmap showing what is planned for the future \citep{Yo2021,
ThielEtAl2020, vanGompelEtAl2016}.  We did see this information mentioned in
some repos, but we did not observe a specific artifact devoted to this purpose.
\item A code of conduct to explicitly say how developers should treat one
another \citep{Yo2021, ThielEtAl2020}.  We did not observe a policy of this
sort, possibly because the number of developers for each MI project is small.
\item Programming style guidelines so that programming labels and formatting are
consistent \citep{TobiasEtAl2018, ThielEtAl2020, OrvizEtAl2017, Zadka2018,
vanGompelEtAl2016, WilsonEtAl2014}.  We saw this information as part of
some developer guides, but only rarely.
\item Checklists can be used in projects to ensure that best practices are
followed by all developers.  Some examples include checklists merging branches
into master \citep{Brown2015}, checklists for saving and sharing changes to the
project \citep{WilsonEtAl2016}, checklists for new and departing team members
\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
releases \citep{HerouxEtAl2008} and checklists for overall software quality
\citep{ThielEtAl2020, SSI2022}.  For MI software, ESPResSo has a checklist for
managing releases, but otherwise they are used only rarely for MI.
\item Uninstall instructions \citep{vanGompelEtAl2016} were not observed for any
of the MI projects.
\end{itemize}

The above discussion shows that, taken together, MI projects fall short of
recommended best practices for research software.  However, MI software is not
alone in this.  Many, if not most, research projects fall short of best
practices.  Software requirements documentation provides a good example of the
common deviation between recommendations and practice.  Although requirements
documentation is recommended by some \citep{TobiasEtAl2018, HerouxEtAl2008,
SmithAndKoothoor2016}, in practice research software developers often do not
produce a proper requirements specification \citep{HeatonAndCarver2015}.
\citet{SandersAndKelly2008} interviewed 16 scientists from 10 disciplines and
found that none of the scientists created requirements specifications, unless
regulations in their field mandated such a document. \citet{Nguyen-HoanEtAl2010}
showed requirements are the least commonly produced type of documentation for
research software in general. When looking at the pain points for research
software developers, \citet{WieseEtAl2019} found that software requirements and
management is the software engineering discipline that most hurts scientific
developers, accounting for 23\% of the technical problems reported by study
participants.  The lack of support for requirements is likely due to the
perception that up-front requirements are impossible for research software
\citep{CarverEtAl2007,SegalAndMorris2008}, but when the instance on ``up-front''
requirements is dropped, allowing the requirements to be written iteratively and
incrementally, requirements are feasible \citep{Smith2016}.

@@

@@

>>>>

As part of filling in the measurement template (from
Section~\ref{sec_grading_software}) we summarized the artifacts observed in each
MI package. Table~\ref{artifactspresent} groups the artifacts by frequency into
categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19 (33-67\%)
packages), and rare (1 to 9 ($<$33\%) packages). The full measurements are
summarized in \citet{Dong2021-Data}.  The details on which projects use which
types of artifacts are summarized in Tables~\ref{tab_maintainability_docs}
and~\ref{tab_Visibility/Transparency_docs} for documents related to
maintainability and visibility, respectively.

\begin{table}[ht!]
    \begin{center}
    \begin{tabular}{ p{4.6 cm} p{5.6 cm} p{5 cm}}
    \toprule
    \textbf{Common} & \textbf{Uncommon} & \textbf{Rare} \\
    \midrule
    README (29) & Build scripts (18) & Getting Started (9)\\
    Version control (29) & Tutorials (18) & Developer's manual (8)\\
    License (28) & Installation guide (16) & Contributing (8)\\
    Issue tracker (28) & Test cases (15) & API documentation (7)\\
    User manual (22) & Authors (14) & Dependency list (7)\\
    Release info. (22) & Frequently Asked Questions (FAQ) (14) & Troubleshooting guide (6)\\
     & Acknowledgements (12) & Product roadmap (5)\\
     & Changelog (12) & Design documentation (5)\\
     & Citation (11) & Code style guide (3)\\
     & & Code of conduct (1)\\
    \bottomrule
    \end{tabular}
    \caption{Artifacts Present in MI Packages, Classified by Frequency (The number 
    in brackets is the number of occurrences)}
    \label{artifactspresent}
    \end{center}
\end{table}

We answer~\rqref{RQ_CompareArtifacts} by comparing the artifacts that we
observed in MI repositories to those observed and recommended for research
software in general. Our comparison may point out areas where some MI software
packages fall short of current best practices. This is not intended to be a
criticism of any existing packages, especially since in practice not every
project needs to achieve the highest possible quality. However, rather than
delve into the nuances of which software can justify compromising which
practices we will write our comparison under the ideal assumption that every
project has sufficient resources to match best practices.

Table~\ref{Tbl_Guidelines} (based on data from \citep{SmithAndMichalski2022})
shows that MI artifacts generally match the recommendations found in nine
current research software development guidelines:
\begin{itemize}
\item United States Geological Survey Software Planning Checklist
\citep{USGS2019},
\item DLR (German Aerospace Centre) Software Engineering Guidelines
\citep{TobiasEtAl2018}, 
\item Scottish Covid-19 Response Consortium Software Checklist
\citep{BrettEtAl2021},
\item Good Enough Practices in Scientific Computing \citep{WilsonEtAl2016},
\item xSDK (Extreme-scale Scientific Software Development Kit) Community Package
Policies \citep{SmithAndRoscoe2018},
\item Trilinos Developers Guide \citep{HerouxEtAl2008},
\item EURISE (European Research Infrastructure Software Engineers') Network
Technical Reference \citep{ThielEtAl2020},
\item CLARIAH (Common Lab Research Infrastructure for the Arts and Humanities)
Guidelines for Software Quality \citep{vanGompelEtAl2016}, and
\item A Set of Common Software Quality Assurance Baseline Criteria for Research
Projects \citep{OrvizEtAl2017}.
\end{itemize}

In Table~\ref{Tbl_Guidelines} each row corresponds to an artifact.  For a given
row, a checkmark in one of the columns means that the corresponding guideline
recommends this artifact.  The last column shows whether the artifact appears in
the measured set of MI software, either not at all (blank), commonly (C),
uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
artifact consistently between guidelines and specific MI software, but the
terminology and the contents of artifacts are not standardized.  The challenge
even exists for the ubiquitous README file.  As illustrated by
\citet{PranaEtAl2018}, the content of README files shows significant variation
between projects.  Although some content is reasonably consistent, with 97\% of
README files contain at least one section describing the `What` of the
repository and 89\% offering some `How` content, other categories are more
variable.  For instance, information on `Contribution`, `Why`', and `Who`,
appear in 28\%, 26\% and 53\% of the analyzed files, respectively
\citep{PranaEtAl2018}.

The frequency of checkmarks in Table~\ref{Tbl_Guidelines} indicates the
popularity of recommending a given artifact, but it does not imply that the most
commonly recommended artifacts are the most important artifacts. Just because an
artifact is not explicitly recommended in a given guidelines, does not mean that
the artifact is not valued by the guideline authors.  They may have exluded it
because it is out of the scope of their recommendations, or outside of their
experience.  For instance, an artifact related to uninstall is only explicitly
mentioned by \citep{vanGompelEtAl2016}), but other guideline authors would
likely see its value.  They may simply feel that uninstall is implied by
install, or they may have never asked themselves whether separate uninstall
instructions are needed.

@@

Two of the items that appear in Table~\ref{artifactspresent} do not appear in
the software development guidelines shown in Table~\ref{Tbl_Guidelines}:
Troubleshooting guide and Developer's manual.  Although these two artifacts
aren't specifically named in the guidelines that we found, the information
contained within them overlaps with the recommended artifacts.  A
Troubleshooting guideline contains information that would typically be found in
a User manual.  A Developer's guide overlaps with information from the README,
INSTALL, Uninstall, Dependency List, Release Information, API documentation and
Design documentation.  In our current analysis, we have identified artifacts by
the names given by the software guidelines and MI examples.  In the future, a
more in-depth analysis would look at the knowledge fragments that are captured
in the artifacts, rather than focusing on the names of the files that collect
these fragments together.

Although the MI community shows examples of 88\% (23 of 26) of the practices we
found in research software guidelines (Table~\ref{Tbl_Guidelines}), three
recommended artifacts were not observed: i) Unistall, ii) Test plans, and iii)
Requirements.  Uninstall is likely an omission caused by the focus on installing
software. Given the storage capacity of current hardware, developers and users
do not generally concerned with uninstall. Moreover, as mentioned above,
uninstall is not particularly emphasized in existing recommendations.  Test
plans were not observed for MI software, but that doesn't mean they weren't
created; it means that the plans are not under version control.  Test plans
would have to at least be implicitly created, since test cases were observed
with reasonable frequency for MI software (uncommon).

MI software is like other research software in its neglect of requirements
documentation.  Although requirements documentation is recommended by some
\citep{TobiasEtAl2018, HerouxEtAl2008, SmithAndKoothoor2016}, in practice
research software developers often do not produce a proper requirements
specification \citep{HeatonAndCarver2015}. \citet{SandersAndKelly2008}
interviewed 16 scientists from 10 disciplines and found that none of the
scientists created requirements specifications, unless regulations in their
field mandated such a document. \citet{Nguyen-HoanEtAl2010} showed requirements
are the least commonly produced type of documentation for research software in
general. When looking at the pain points for research software developers,
\citet{WieseEtAl2019} found that software requirements and management is the
software engineering discipline that most hurts scientific developers,
accounting for 23\% of the technical problems reported by study participants.
The lack of support for requirements is likely due to the perception that
up-front requirements are impossible for research software
\citep{CarverEtAl2007,SegalAndMorris2008}, but when the instance on ``up-front''
requirements is dropped, allowing the requirements to be written iteratively and
incrementally, requirements are feasible \citep{Smith2016}.

Table~\ref{Tbl_Guidelines} shows several artifacts that are rarely observed in
practice.  A theme amount these rare artifacts is that many of them are related
to developers more than users.  For instance, the rare artifacts include
Contributing, developer Code of conduct, Code Style Guide, Product roadmaps,
Requirements, Design documentation, API documentation and a Test plan.  The rare
artifacts for MI software are similar to the rare artifacts for Lattice
Boltzmann solvers \citep{Michalski2021}.  However, MI software differs from
ocean modelling software, since, according to \citep{JungEtAl2022}, the later
pays relatively less attention to testing.

To improve MI software in the future, an increased use of checklists could help.
Checklists can be used in projects to ensure that best practices are followed by
all developers.  Some examples include checklists merging branches into master
\citep{Brown2015}, checklists for saving and sharing changes to the project
\citep{WilsonEtAl2016}, checklists for new and departing team members
\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
releases \citep{HerouxEtAl2008} and checklists for overall software quality
\citep{ThielEtAl2020, SSI2022}.  For instace, for Lattice Boltzmann solver
software, ESPResSo has a checklist for managing releases \citep{Michalski2021}.

The above discussion shows that, taken together, MI projects fall somewhat short
of recommended best practices for research software.  However, MI software is
not alone in this.  Many, if not most, research projects fall short of best
practices.  A gap exists in scientific computing development practices and
software engineering recommendations \citep{Storer2017, Kelly2007,
OwojaiyeEtAl2021_CSE}. \citet{JohansonAndHasselbring2018} observe that the
state-of-the-practice for SCS in industry and academia does not incorporate
state-of-the-art SE tools and methods.  This causes sustainability and
reliability problems \citep{FaulkEtAl2009}. Rather than benefit from capturing
and reusing previous knowledge, projects waste time and energy ``reinventing the
wheel'' \citep{deSouzaEtAl2019}.

@@

\wss{observed artifacts (Section~\ref{Sec_CompareArtifacts}) - human judgement,
not always given the expected name, but the content is there.}

@@

\wss{Talk about increasing the number of developers.  CONTRIBUTING is rare
(Table~\ref{artifactspresent}), but it doesn't have to be.  Related to this
information for new developers is rare: Code Style Guide, Requirements, Design
Doc, API.}
